---
title: "<center>Answers to the exercises from Chapter 2<center>"
author: "<center>Sébastien Wieckowski<center><br>"
date: "<center> _`r Sys.Date()`_ <center>"
output:
    html_document:
        code_folding: show
        df_print: default
        highlight: tango
        number_sections: yes
        theme: simplex
        toc: yes
        toc_float:
            collapsed: yes
            smooth_scroll: no
    pdf_document:
        fig_caption: yes
        latex_engine: xelatex
        toc: yes
geometry: margin=1.5cm
---

```{r, include=FALSE}
library(rmarkdown)
library(tinytex)
library(knitr)
```

# Data exercises

## Question 1

Create a character vector called my_names that contains all your first, middle and last names as elements. Calculate the length of my_names.

```{r, echo = TRUE}
my_names <- c("Sébastien", "Wieckowski")
length(my_names)
```

## Question 2

Create a second numeric vector called which which corresponds to my_names. The entries should be the position of each name in the order of your full name. Verify that it has the same length as my_names.

```{r, echo = TRUE}
which <- c(1,2)
length(which)
```

We verify if the dimensions of the two vectors are identical: `r length(my_names) == length(which)`.

## Question 3

Create a dataframe called names, which consists of the two vectors my_names and which as columns. Calculate the dimensions of names.

```{r, echo = TRUE}
names <- data.frame(my_names, which)
```

The dimension of the vector `names` is `r dim(names)`.

## Question 4

Create a new dataframe new_names with the which column converted to character type. Verify that your command worked using str().

```{r, echo = TRUE}
new_names <- data.frame(my_names, as.character(which))
str(new_names)
```

***

## Question 5

Load the ugtests data set via the peopleanalyticsdata package or download it from the internet7. Calculate the dimensions of ugtests and view the first three rows only.

```{r, echo = TRUE}
ugtests <- read.csv("http://peopleanalytics-regression-book.org/data/ugtests.csv")
dim(ugtests)
head(x=ugtests, n=3)
```

## Question 6

View a statistical summary of all of the columns of ugtests. Determine if there are any missing values.

```{r, echo = TRUE}
summary(ugtests)
```

There are `r sum(is.na(ugtests))` missing values.

## Question 7

View the subset of ugtests for values of Yr1 greater than 50.

```{r, echo = TRUE}
ugtests %>%
  subset(subset=(ugtests$Yr1 > 50)) %>%
  head(n = 10)
```

## Question 8

Install and load the package dplyr. Look up the help for the filter() function in this package and try to use it to repeat the task in the previous question.

```{r, echo = TRUE}
library(dplyr)
# help(filter) # look up at the help

ugtests %>%
  filter(Yr1 > 50) %>%
  head(n = 10)
```

## Question 9

Write code to find the mean of the Yr1 test scores for all those who achieved Yr3 test scores greater than 100. Round this mean to the nearest integer.

### R base-style

```{r, echo = TRUE}
ugtests$Yr1 %>%
  subset(ugtests$Yr3 > 100) %>%
  mean() %>%
  round()
```

### Attempt using Dplyr

```{r, echo = TRUE}
ugtests %>%
  filter(Yr3 > 100) %>%
  summarise(avg=mean(Yr1)) %>% # mean function doesn't work (see below)...
  round()
```

The `mean` base function doesn't work on the output of the pipe because the upstream [`filter` function generates a data frame structure and not a vector](https://stackoverflow.com/questions/49844902/difference-between-pull-and-select-in-dplyr).
One solution is to use the `colMeans` function:

```{r, echo = TRUE}
ugtests %>%
  filter(Yr3 > 100) %>%
  select(Yr1) %>%
  colMeans() %>%
  round()
```

### Question 10

Familiarize yourself with the two functions filter() and pull() from dplyr. Use these functions to try to do the same calculation in the previous question using a single unbroken piped command. Be sure to namespace where necessary.

```{r, echo = TRUE}
ugtests %>%
  filter(Yr3 > 100) %>%
  pull(Yr1) %>%
# pull() creates a vector -- which, in this case, is numeric --
# whereas select() creates a data frame. Basically, pull() is the equivalent
# to writing ugtests$Yr1 or ugtests[, "Yr1"], whereas select() removes all of
# the columns except for Yr1 but maintains the data frame structure.
  mean() %>%
  round()
```

***

## Question 11

Create a scatter plot using the ugtests data with Final scores on the y axis and Yr3 scores on the x axis.

```{r, echo = TRUE, fig.cap="Scatter plot.", fig.width=7, fig.height=5}
plot(
  x = ugtests$Yr3,
  y = ugtests$Final,
  xlab = "Yr3 scores",
  ylab = "Final scores",
  main = "ugtests scatter plot"
)
```

## Question 12

Create your own 5-level grading logic and use it to create a new finalgrade column in the ugtests data set with grades 1-5 of increasing attainment based on the Final score in ugtests. Generate a histogram of this finalgrade column.

```{r, echo = TRUE}
ugtests$final_grade <- ugtests$Final %>%
  cut(breaks = 5, ordered_result = T, labels = c(1:5)) %>%
  as.numeric()

head(ugtests)
```

```{r, echo = TRUE, fig.cap="Histogram.", fig.width=5, fig.height=4}
hist(ugtests$final_grade, breaks = 0:5)
```

## Question 13

```{r, echo = TRUE, fig.cap="Boxplot.", fig.width=6, fig.height=4}
boxplot(
  formula = Yr3 ~ final_grade,
  data = ugtests,
  xlab = "Final score grade",
  ylab = "Yr3 score",
  main = "box plot"
)
```

***

## Session info

The output from running `sessionInfo` is shown below and details all packages and version necessary to reproduce the results from this report.

```{r session_info}
sessionInfo()

# for compiling the Rmd document to HTML and PDF using radian:
# rmarkdown::render("chapter2.Rmd", 'html_document')
# rmarkdown::render("chapter2.Rmd", 'pdf_document')
```
